Autopolyploid establishment through polygenic adaptation

### 2025-03-26 doc/paper/response.md

## Associate Editor

>Thank you for submitting your article to Evolution. It has now received two
reviews, both of which recognize its valuable contributions and essentially
recommend publication after relatively minor revision. I share these views.
While both reviews are positive, they make a number of worthwhile suggestions
for you to consider for revising the manuscript. One reviewer has suggested
combining the Results and the Discussion into a single section to avoid
unnecessary discussion of the results in the current Results section.
Personally, I would favor retaining both sections, **moving more discussive text
from the Results to the Discussion**. I leave the decision to you, but I do agree
with the reviewers that some resorting of the Results text is needed. I also
agree that the **Introduction could be better motivated by referring more to
empirical work**, and the **Discussion section could relate the results more to
other theoretical findings** in the literature. Overall, think that revision will
be relatively straightforward, but I would ask that you consider all the
suggestions made by the reviewers and, if you disagree with them, that you
justify and explain your reasons in a response letter.

I have now revised the Introduction slightly to include more discussion of
empirical work, although, in the interest of brevity (the manuscript is already
near the Evolution word limit), I mostly resorted to explicitly referring the
reader to the article by @griswold2021. In the latter, the author provides a
nice overview of relevant recent studies.

Regarding the Results and the Discussion sections: I do not want to do away
with Discussion section altogether (as suggested by Reviewer 1), given that I
wish to discuss factors relevant to the problem which are not studied in my
article (dominance, inbreeding depression, *etc.*) and want to relate my work
to other recent theoretical work.
I believe it is better to separate the latter topics from the description *and*
interpretation of the actual results. 
I have sought to move 'more discussive text' to the Discussion section,
although I found there was fairly limited scope to do so when I seek to keep
the interpretation of the results more or less self-contained. **TODO!**

@@

>A small concluding section where the authors mentioned other questions to
study with their model will be useful in my opinion. The author proposed to
complexify the current model of adaptation with dominance or to relax some
hypotheses, but I think other biological scenarios could be studied too.

**TODO**

>Defining establishment when N=100: Is it a classic way to define establishment
in such models?  Or does it come from your experience with the model and you
see that Under N=100 extinction remains likely? More details would be needed
here.

$N=100$ is indeed an arbitrary choice. This threshold was also adopted in
@barton2018, so it makes comparisons of establishment probabilities *etc.* with
their paper somewhat more straightforward.
$N=100$ seems to be a reasonable threshold in that extinction becomes
sufficiently unlikely, while we do not have to simulate large populations (our
approach requires tracking the $N \times N$ matrix of identity coefficients). 
I did a check by putting the threshold at $N=150$, and the establishment
probabilities do not change appreciably.

@@

>Abstract: I am not sure the reference to Barton et al. (2017) is appropriate
here. Maybe a more general definition would fit better.

The reason for the reference is that there appears to be some confusion about
different models that bear the name 'infinitesimal model' (as outlined in for
instance @turelli2017 and @walsh2018). I wanted to make it clear that we use
the "Gaussian descendants" infinitesimal model, which is very carefully
outlined in @barton2017. However, I guess this is sufficiently clear from the
introduction and the methods sections, so the reference in the abstract can
indeed be omitted, as it is in the revised manuscript.

@@

>Discussion about inbreeding depression: again, some empirical data are
available to support some statements (Husband et al. (2008), Clo & Kolar
(2022)).

I have added the suggested references.

## Reviewer 2

>This paper is an important foundational work in the area of theoretical polyploid
evolutionary genetics. It presents an infinitesimal model of trait evolution for
triploids and tetraploids, and then applies the model to assess conditions under
which a tetraploid population will establish in a peripheral region connected to
a central core by dispersal. In terms of the infinitesimal model, several aspects
are very helpful

- Demonstrating how genetic effects are scaled across triploids and tetraploids
such that segregating variance is equivalent across ploidies, including with diploids
- How to combine the genetic effects in an individual with mixed-ploidy ancestry
- Calculating inbreeding and co-ancestry values for mixed-ploidy individuals
- Calculating the effect of unreduced gamete formation on segregation variance
across diploids, triploids and tetraploids

Other informative analyses include recursive or Markov models of null expecta-
tions for frequencies of ploidies, times to diploid ancestry and effective popula-
tion size across a population of mixed-ploidy.
Besides this set of results, the paper then examines conditions under which a
tetraploid population is established in a peripheral population connected to a
central core population that is predominantly composed of diploids. Given the
additive nature of gene-action in the infinitesimal model, this analysis will likely
be of interest to empiricists because additive effects are thought to be the least
constrained basis for something to evolve and there is empirical evidence that
the genetic basis of polyploid establishment in some species is additive. This is
in contrast to the theoretical results of Griswold (2021), which highlighted the
potential for a recessive-basis for autotetraploid establishment in a peripheral
region.
Main comments focus on mutation and adding background/context.

@@

I have not considered mutation at all so far because I assumed it would not be
relevant for the population sizes and timescales considered in my study.
Specifically, any individual at the time of establishment derives from a
completely outbred migrant individual a relatively short time in the past (10
to 100 generations, say), so the contribution of new mutation to differences in
establishment probability between diploids and tetraploids should be
negligible.
This is definitely the case when there is ongoing migration: as long as $m$ is
sufficiently greater than $\mu$ and $V_m$ is sufficiently smaller than the
genetic variance in the mainland population, we will have that for every
mutation contributing a very slight decrease in the loss of segregation
variance on the island, there will be many more migrant arrivals, which each
introduce a completely unrelated genotype.

Of course, when considering the mixed-ploidy infinitesimal model independently
of the establishment question, mutation does become a relevant topic. I mention
this in the revised discussion but I think it is out of scope to discuss this
in any detail in the paper. I think that my approach for scaling across ploidy
levels that is explicitly based on underlying allelic effects in a finite-locus
model extends readily to mutation. The main difference between ploidy levels is
of course that the mutational target size is proportional to the ploidy level,
so that, unless the mutational variance is exactly halved in tetraploids, the
contribution of mutation to genetic variance ($\mu V_m$) in tetraploids is
larger than in diploids.

>In the intro providing empirical examples of polyploid establishment in
peripheral habitats would broaden readership and connections to the
literature. Also, reviewing empirical work on the genetics of adaptation to
peripheral environments in polyploids would broaden connections to the
literature.

I agree with the relevance of empirical work on polyploid establishment and
adaptation to marginal habitats for the introduction of the paper. However, I
think that to present an actual review of empirical studies here is a bit out
of scope. Griswold presents a brief review in his 2021 paper, and I prefer to
refer to his paper instead of paraphrasing his paragraph. Specifically, I now
write:
"Many empirical studies of mixed-ploidy populations show a pattern where
polyploids established in peripheral habitats at the edge of a species’ range
(reviewed in Griswold (2021)), and these observations for specific study
systems are in accord with large scale biogeographical patterns (Rice et al.,
2019)."

With regard to adaptation to peripheral habitats: I refer to an important (if
somewhat old) review article of Kawecki (2008) and recent theoretical work by
Sachdeva et al. I think this gives a good sample of where to look for more
references.
When it comes to empirical work on the *genetics of adaptation to peripheral
environments in polyploids*, I am quite frankly not aware of relevant studies...
Reviewing empirical work on adaptation to peripheral habitats in general is
somewhat out of scope, as this is a vast field on its own with which I myself
am not that well acquainted, and the paper is already at the word limit for
Evolution research articles.

@@

>The discussion could relate findings to other theoretical works more. For ex-
ample, it does not go back and place the results in the context of Oswald and
Nuismer (2011), which assumed a two-locus model and varied dosage effects, as
well as competition between diploids and tetraploids, including no competition.
In addition it allowed for assortative mating.

**TODO** (Discussion)

>It would be informative to expand on the the consequences of different scalings
of genetic effects. This paper made the sensible choice to scale such that segre-
gation variance is equal between ploidies as a baseline. Griswold (2021) scaled
genetic effects such that autotetraploid and diploid individuals have equal fit-
ness as a function of proportional allele count in an individual. A consequence
of Griswold’s scaling is autotetraploids have lower additive genetic variance at
HWE (Gallais 2003, p. 185), so are expected to respond more slowly in the
absence of another process. Gallais (2003, p. 186) mentions some studies that
compared additive genetic variance between diploids and autotetraploids.

The consequences of different scalings is taken up in the results and
discussion throughout the paper where I believe this is relevant.
In the methods section the relationship between allelic effects and additive
genetic variance is stated explicitly.
In the results shown in Fig. 2, the effect of scaling assumptions on
establishment probability is explicitly shown and described.
In the section on establishment with recurrent migration, the phenomenon that
polyploid offspring is more extreme on average when allelic effects are not
exactly scaled by one half (as in Griswold) is taken up. 
I now added some further discussion, referring explicitly to Griswold's scaling
in the Discussion section. I now also refer to @porturas2019. Sadly, I
currently do not have access to Gallais' book.

@@

I am unsure whether this is a good idea, as it may lead to some confusion as to
which version of the infinitesimal model we are working with.
The model described in Bulmer (1971) is not exactly equivalent to the Gaussian
descendants approximation that I use (as outlined in Turelli (2017), which I
cite).
If the reference is suggested by the reviewer in order to cite a more
'original' reference, I don't think there is a reason to single out Bulmer
(1971), one could refer to Fisher (1918) or work by Alan Robertson in the
1960's.
I think the reference to Turelli (2017) which reviews the history of the
infinitesimal model suffices, and prevents confusion as to what we mean when we
refer to 'the infinitesimal model'.

>p.7: Suggest adding “, across gene copy arrangements and ploidies” at the end
of the last sentence before the “Establishment” section.

I do not fully understand this suggestion. I personally find this makes the
sentence more confusing, as I never talk of gene copy arrangements.

>p. 7, Establishment section: The organization of paragraphs two and three in
this section is a bit confusing. N is used before it is formally defined. In my
first reading I thought there was migration and then selection, whereas there is
selection and then migration. It would also help support a self-contained paper
to give the expression for $E[e\gamma(...) ]$.

### 2025-03-26 doc/paper/paper.tex
revising

Consider for instance a randomly mating diploid population.
Under the commonly accepted view that new polyploid plants are mostly formed
through the union of unreduced gametes
\citep{bretagnolle1995,herben2016,kreiner2017b}, a new tetraploid individual
originating by a chance encounter of two unreduced diploid gametes (an event
occurring at an appreciable rate; \cite{kreiner2017}) is highly unlikely to
establish a stable tetraploid subpopulation, as most of its gametes will end up
in unfit hybrids of odd ploidy level (a phenomenon referred to as `triploid
block', see \cite{ramsey1998,kohler2010,brown2024}).
This positive frequency dependence effect is commonly referred to as
\textit{minority cytotype exclusion} (MCE), after \cite{levin1975}.
It is well-appreciated that, as a consequence of MCE, the rate of unreduced
gamete formation needs to be extraordinarily high for tetraploids to establish
in a large random mating initially diploid population (\cite{felber1997}, see
also \cref{s-sec:det}).

Hence, to explain the widespread occurrence of mixed-ploidy populations,
additional factors besides the continuous formation of polyploids through the
union of unreduced gametes need to be considered.
Firstly, chance establishment of tetraploids through drift could occur.
Indeed, the problem is somewhat analogous to the spread of underdominant
chromosomal rearrangements, where local establishment through genetic drift and
subsequent spreading in a subdivided population through local extinction and 
recolonization dynamics has been suggested as a plausible model
\citep{lande1985}.
However, MCE is quite strong in randomly mating mixed-ploidy populations, and
the population size has to be very small for local tetraploid establishment to
occur at an appreciable rate (\cite{rausch2005}, see also \cref{s-sec:mchain}).
Secondly, any form of \textit{prezygotic isolation} between cytotypes could
promote establishment of polyploid cytotypes by alleviating MCE.
Particularly relevant are assortative mating by cytotype  (for instance through
phenological differences across cytotypes, or differences in pollinators;
\cite{kolar2017}), self-fertilization \citep{rausch2005,novikova2023}, 
localized dispersal \citep{baack2005,kolar2017} and asexual reproduction
\citep{vandrunen2022}.
Finally, selection may be invoked to explain the establishment of polyploids.
Tetraploids may have higher relative fitness than their diploid counterparts
due to reduced inbreeding depression
\citep{ronfort1999,otto2000,husband2008,clo2022b}, or due to being
better adapted to (changing) environmental conditions \citep{vandepeer2021}.

None of these factors is likely to explain by itself the establishment of
polyploids, and the consensus in the field appears to be that some mix of the
above is required to explain the occurrence of mixed-ploidy populations in
nature \citep{kolar2017,mortier2024}.
Many empirical studies show a pattern where polyploids established in
peripheral habitats at the edge of a species' range (reviewed in
\cite{griswold2021}), in accordance with large-scale biogeographical patterns
\citep{rice2019}.
In a peripheral habitat, a new polyploid population may be more likely to reach
an appreciable size as it may evade to some extent the negative effects of MCE.
However, at the same time, such peripheral habitats are likely to present
adaptive challenges to establishment
\citep{kawecki2008adaptation,sachdeva2022b}, and if polyploids are able to
colonize such habitats at an appreciable rate, they must somehow be better
adapted to local conditions, or more able to adapt to those conditions despite
inbreeding and maladaptive migration, compared to diploids.
Indeed, when the source population is dominated by diploids, the probability
that a migrant individual is tetraploid will be very small ($\sim u^2$, where
$u$ is the probability that a diploid produces an unreduced gamete in meiosis),
so that tetraploids need a substantial advantage during colonization if they
are to establish before diploids do.

While there have been substantial modeling efforts aimed at understanding
autotetraploid establishment within diploid populations \citep{levin1975,
felber1991, felber1997, rausch2005, oswald2011, clo2022c}, the problem of
polyploid establishment in peripheral habitats remains largely unaddressed
(but see \cite{griswold2021}),
despite its centrality to verbal arguments about the establishment of
polyploids in natural populations \citep{kolar2017, vandepeer2021, clo2022d}.

@@

\begin{table}[t]
\caption{Glossary of the notation used in the main text.
} \label{tbl:glossary}
\centering
\small
\begin{tabularx}{\linewidth}{lX}
\cline{1-2}
\textbf{notation}   & \textbf{description}   \\ \cline{1-2}
$N$ & total population size\\
$N_k$ & population size of the $k$-ploid cytotype \\
$\pi_k$ & deterministic equilibrium frequency of the $k$-ploid cytotype \\
$u$ & probability of unreduced gamete formation ($u=u_{22}=1-u_{42}$)\\
$v$ & probability that a triploid produces a haploid/diploid gamete
  ($v=u_{31}=u_{32}$)\\
$m$ & expected number of migrants per generation arriving in the new habitat \\
$z_i$ & trait value of individual $i$ \\
$c_i$ & ploidy level of individual $i$ \\
$g_i$ & ploidy level of gamete produced by individual $i$ in a particular cross\\
$V$ & segregation variance in the reference diploid population \\
$V_{i,k}$ & gametic segregation variance associated with the production of a
  $k$-ploid gamete by individual $i$ \\
$\Vx_k$ & genetic variance associated with a haploid genome in the $k$-ploid
  reference population (i.e. a $k$-ploid non-inbred population at HWLE) \\
$\beta_{k}$ & scaling factor for allelic effects in $k$-ploids \\
$F_i$ & inbreeding coefficient in individual $i$ \\
$\Phi_{ij}$ & coancestry coefficient for individuals $i$ and $j$ \\
$\alpha_k$ & probability that the two genes at a locus in a diploid gamete
  formed by a $k$-ploid individual descend from the same parental gene copy\\
$\gamma$ & strength of directional selection in the new habitat\\
$\theta$ & trait value beyond which the growth rate becomes positive in the new
habitat \\ $w_{ij}$ & fitness of parental pair $(i,j)$ \\
$w_{ij}^{kl}$ & expected fitness of offspring from parental pair $(i,j)$ when
$i$ contributes a $k$-ploid gamete and $j$ contributes a $l$-ploid gamete \\
$\sigma_k$ & rate of self-fertilization in $k$-ploids \\
$\rho_k$ & probability of assortative mating in $k$-ploids \\
\cline{1-2}
\end{tabularx}%
\end{table}

@@

The observation that polyploid populations tend to inhabit more extreme
habitats or occur at the edge of the range of their conspecific diploids
has spurred considerable interest among botanists and evolutionary biologists
\citep{rice2015,kolar2017,vandepeer2021,griswold2021,mortier2024}.
An important question is whether such patterns emerge because polyploids are
somehow more tolerant to extreme environmental conditions (i.e. they somehow
are intrinsically more fit than diploids in marginal habitats), or whether other
aspects of the population dynamics of mixed-ploidy populations may favor the
establishment of polyploid subpopulations.

@@

Importantly, we assume no intrinsic advantage or disadvantage of polyploids in
the marginal habitat, i.e. the expected fitness of a migrant individual is the
same regardless of the ploidy level.
Differences in establishment probabilities are hence caused solely by aspects
of autopolyploid genetics and the barrier to gene flow between subpopulations
of different ploidy levels.
This is undoubtedly unrealistic.
For instance, neopolyploids are likely to suffer intrinsic fertility issues due
to meiotic irregularities associated with multivalent formation
\citep{bomblies2016,novikova2023}, and triploids may be inviable due to
issues with endosperm development \citep{bretagnolle1995}.
Similarly implausible is the assumption of a constant equilibrium genetic
variance across cytotypes ($\beta_4=\sqrt{1/2}$ in our model), which we used in
most of our results.
It would be very interesting to experimentally assess whether typical
quantitative traits in mixed-ploidy populations follow approximately the
mixed-ploidy infinitesimal model we outlined in this paper, and to actually
estimate how the variance scales across ploidy levels.
In addition, we have assumed a relatively high and constant rate of unreduced
gamete formation $u$ and triploid fertility $v$ in all our simulations (5\%),
whereas these are known to be variable across the population, and at least in
part genetically determined \citep{kreiner2017,clo2022c}.
We ignore such complications, and hence do not take the actual establishment
probabilities very serious, focusing instead on how migration load and
prezygotic isolation affect the tetraploid establishment probability.

>>>>

Consider for instance a randomly mating diploid population.
Under the commonly accepted view that new polyploid plants are mostly formed
through the union of unreduced gametes
\citep{bretagnolle1995,herben2016,kreiner2017b}, a new tetraploid individual
originating by a chance encounter of two unreduced diploid gametes (an event
occurring at an appreciable rate; \cite{kreiner2017}) is highly unlikely to
establish a stable tetraploid subpopulation, as most of its gametes will end up
in unfit hybrids of odd ploidy level (a phenomenon referred to as `triploid
block', see \cite{ramsey1998,kohler2010,brown2024}).
This positive frequency dependence effect in mixed-ploidy populations is
commonly referred to as \textit{minority cytotype exclusion} (MCE), after
\cite{levin1975}.
It is well-appreciated that, as a consequence of MCE, the rate of unreduced
gamete formation needs to be extraordinarily high for tetraploids to establish
a stable subpopulation in a large random mating population initially dominated
by diploids (\cite{felber1997}, see also \cref{s-sec:det}).

Hence, to explain the widespread occurrence of mixed-ploidy populations,
additional factors besides the continuous formation of polyploids through the
union of unreduced gametes need to be considered.
Firstly, chance establishment of tetraploids through drift could occur.
Indeed, the problem is somewhat analogous to the spread of underdominant
chromosomal rearrangements, where local establishment through genetic drift and
subsequent spreading in a subdivided population by means of local extinction
and recolonization has been suggested as a plausible model \citep{lande1985}.
However, MCE is quite strong under random mating, and the population size has
to be very small for local tetraploid establishment to occur at an appreciable
rate (\cite{rausch2005}, see also \cref{s-sec:mchain}).
Secondly, any form of \textit{prezygotic isolation} between cytotypes could
promote establishment of polyploid cytotypes by alleviating MCE.
Particularly relevant are assortative mating by cytotype  (for instance through
phenological differences across cytotypes, or differences in pollinators;
\cite{kolar2017}), self-fertilization \citep{rausch2005,novikova2023}, 
localized dispersal \citep{baack2005,kolar2017} and asexual reproduction
\citep{vandrunen2022}.
Finally, selection may be invoked to explain the establishment of polyploids.
Tetraploids may have higher relative fitness than their diploid counterparts
due to reduced inbreeding depression
\citep{ronfort1999,otto2000,husband2008,clo2022b}, or due to being
better adapted to (changing) environmental conditions \citep{vandepeer2021}. 
However, none of these factors is likely to explain by itself the establishment
of polyploids, and the consensus in the field appears to be that some mix of
the above is required to explain the occurrence of mixed-ploidy populations in
nature \citep{kolar2017,mortier2024}.

Many empirical studies of mixed-ploidy populations find that polyploids
established in peripheral habitats at the edge of a species' range (reviewed in
\cite{griswold2021}), and this is in accord with large scale biogeographical
patterns \citep{rice2019}.
This raises the question: which aspects of the process of adaptation to marginal
habitats could promote the establishment of polyploid populations?
In a peripheral habitat, a new polyploid population may be more likely to reach
an appreciable size as it evades to some extent the negative effects of MCE.
However, at the same time, such peripheral habitats are likely to present
adaptive challenges to establishment
\citep{kawecki2008adaptation,sachdeva2022b}, and if polyploids are able to
colonize such habitats at an appreciable rate, they must somehow be better
adapted to local conditions, or more able to adapt to those conditions despite
inbreeding and maladaptive migration, compared to diploids.
Indeed, when the source population is dominated by diploids, the probability
that a migrant individual is tetraploid will be very small ($\sim u^2$, where
$u$ is the probability that a diploid produces an unreduced gamete in meiosis),
so that tetraploids need a substantial advantage during colonization if they
are to establish before diploids do.
While there have been substantial modeling efforts aimed at understanding
autotetraploid establishment within diploid populations \citep{levin1975,
felber1991, felber1997, rausch2005, oswald2011, clo2022c}, the problem of
polyploid establishment in peripheral habitats remains largely unaddressed
(but see \cite{griswold2021}),
despite its centrality to verbal arguments about the establishment of
polyploids in natural populations \citep{kolar2017, vandepeer2021, clo2022d}.

@@

\begin{table}[t]
\caption{Glossary of the notation used in the main text.
} \label{tbl:glossary}
\centering
\small
\begin{tabularx}{\linewidth}{lX}
\cline{1-2}
\textbf{notation}   & \textbf{description}   \\ \cline{1-2}
$N$ & total population size\\
$N_k$ & population size of the $k$-ploid cytotype \\
$\pi_k$ & deterministic equilibrium frequency of the $k$-ploid cytotype \\
$u$ & probability of unreduced gamete formation ($u=u_{22}=1-u_{42}$)\\
$v$ & probability that a triploid produces a haploid/diploid gamete
  ($v=u_{31}=u_{32}$)\\
$m$ & expected number of migrants per generation arriving in the new habitat \\
$z_i$ & trait value of individual $i$ \\
$c_i$ & ploidy level of individual $i$ \\
$g_i$ & ploidy level of gamete produced by individual $i$ in a particular cross\\
$V$ & segregation variance in the reference diploid population \\
$V_{i,k}$ & gametic segregation variance associated with the production of a
  $k$-ploid gamete by individual $i$ \\
$\Vx_k$ & genetic variance associated with a haploid genome in the $k$-ploid
  reference population (i.e. a $k$-ploid non-inbred population at HWLE) \\
$\beta_{k}$ & scaling factor for allelic effects in $k$-ploids \\
$F_i$ & inbreeding coefficient in individual $i$ \\
$\Phi_{ij}$ & coancestry coefficient for individuals $i$ and $j$ \\
$\alpha_k$ & probability that the two genes at a locus in a diploid gamete
  formed by a $k$-ploid individual descend from the same parental gene copy\\
$\gamma$ & strength of directional selection in the new habitat\\
$\theta$ & trait value beyond which the growth rate becomes positive in the new
    habitat \\ 
$w_{ij}$ & fitness of parental pair $(i,j)$ (expected fitness of
    offspring of pair $(i,j)$)\\
$w_{ij}^{kl}$ & expected fitness of offspring of parental pair $(i,j)$ when
$i$ contributes a $k$-ploid gamete and $j$ contributes a $l$-ploid gamete \\
$\sigma_k$ & rate of self-fertilization in $k$-ploids \\
$\rho_k$ & probability of assortative mating in $k$-ploids \\
\cline{1-2}
\end{tabularx}%
\end{table}

@@

The observation that polyploid populations tend to inhabit more extreme
habitats or occur at the edge of the range of their conspecific diploids
has spurred considerable interest among botanists and evolutionary biologists
\citep{kolar2017,rice2019,vandepeer2021,griswold2021,mortier2024}.
An important question is whether such patterns emerge because polyploids are
somehow more tolerant to extreme environmental conditions (i.e. they somehow
are intrinsically more fit than diploids in marginal habitats), or whether other
aspects of the population dynamics of mixed-ploidy populations may favor the
establishment of polyploid subpopulations.

@@

Importantly, we assume no intrinsic advantage or disadvantage of polyploids in
the marginal habitat, i.e. the expected fitness of a migrant individual is the
same regardless of the ploidy level.
Differences in establishment probabilities are hence caused solely by aspects
of autopolyploid genetics and the barrier to gene flow between subpopulations
of different ploidy levels.
This is undoubtedly unrealistic.
For instance, neopolyploids are likely to suffer intrinsic fertility issues due
to meiotic irregularities associated with multivalent formation
\citep{bomblies2016,novikova2023}, and triploids may be inviable due to
issues with endosperm development \citep{bretagnolle1995}.

Similarly implausible is the assumption of a constant equilibrium genetic
variance across cytotypes ($\beta_4=\sqrt{1/2}$ in our model), which we used in
most of our results (but see \cref{fig:est1}).
The meta-analysis performed by \cite{porturas2019} suggests that phenotypic
variation across ploidy levels is indeed fairly constant, so this assumption
may not yield too bad an approximation.
It should be noted however that other authors have made different assumptions
on how allelic effects (and hence genetic variance) scale across ploidy levels
(in particular \cite{griswold2021}, who made an assumption equivalent to
$\beta_4 = 1/2$ in our model), and such assumptions evidently impact the
establishment probability (\cref{fig:est1}).
Clearly, it would be very interesting to experimentally assess whether
typical quantitative traits in mixed-ploidy populations follow approximately
the mixed-ploidy infinitesimal model we outlined in this paper, and to actually
estimate how the variance scales across ploidy levels.
In addition, we have assumed a relatively high and constant rate of unreduced
gamete formation $u$ and triploid fertility $v$ in all our simulations (5\%),
whereas these are known to be variable across the population, and at least in
part genetically determined \citep{kreiner2017,clo2022c}.
We ignore such complications, and hence do not take the actual establishment
probabilities very serious, focusing instead on how migration load and
prezygotic isolation affect the tetraploid establishment probability.
